---
title: "Examining Repeatability in Different Rush Situations"
author: "Jason Baik"
date: "4/6/2018"
output: 
  html_document:
    toc: yes
    toc_float: yes
    code_folding: hide
    theme: readable
    highlight: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=2,
                      cache.rebuild = F, 
                      message = F, warning = F,
                      autodep=TRUE,
                      error = TRUE, comment = "")
```

### Key Takeaways

* Its hard to predict the number of shots in two-plus odd-man rushes over a season.

* When there is a one-man advantage, shooting percentages are repeatable.

* All zones of the primary pass show no repeatability in shooting perecentages.

* Rebounds are a random occurence during a hockey game and thus, rebound shooting percentages (percentages of rebound goals scored off of rebounds) are unpredictable.


### Introduction
I will be using the Passing Project Data of the 2015-2016 NHL season organized by [Ryan Stimson](https://twitter.com/RK_Stimp). The data was collected by a group of volunteers who tracked every pass leading to a shot. According to his [MIT Sloan Sports Analytics Conference paper](http://www.sloansportsconference.com/wp-content/uploads/2017/02/1614.pdf)
, "Each tracker was required to complete 2 - 3 training games for validation by the project leader to ensure consistency in classification of passes and recording of data. Occasional spot-checking of games was also conducted. 565 total games were tracked from the 2015-2016 season"

This article is geared for beginners and experts alike as I will be walking through the code chunks by means of comments on top of each code chunk. 

I would like to thank my mentor, Sam Ventura of the Pittsburgh Penguins. Without him, I could not have written this article.

##### The following hidden code imports the data and maniplulates it.

```{r}
# Read and manipulate data
library(tidyverse)
library(ggimage)
library(kableExtra)
```

```{r, warning = FALSE}
pass <- readxl::read_excel("~/Dropbox/passing_project/passing_data_code/passing_data_1516.xlsx") %>% 
  rename(period = Period, 
         time = Time,
         strength = Strength,
         team = Team,
         shooter_number = Shooter,
         shot_type = `Shot Type?`,
         a1 = A1,
         a2 = A2,
         a3 = A3, 
         a1_zone = `A1 Zone`,
         a2_zone = `A2 Zone`,
         a3_zone = `A3 Zone`,
         scoring_chance_pass = `SC?`,
         shot_on_goal = `SOG?`,
         odd_man = Oddman,
         goal = `G?`,
         player_rebound = `RB/2C`,
         rebound_shot_on_goal = `RB/2C SOG?`,
         rebound_goal = `RB/2C G?`,
         home_score = `Home Score State`,
         goalie_number = Goalie,
         game_id = `Game ID`,
         date = Date,
         home_team = `Home:`,
         away_team = `Away:`) %>% 
  mutate(scoring_chance_pass = ifelse(is.na(scoring_chance_pass), 0, 1),
         shot_on_goal = ifelse(is.na(shot_on_goal), 0, 1),
         goal =  ifelse(is.na(goal), 0, 1),
         rebound_shot_on_goal = ifelse(is.na(rebound_shot_on_goal), 0, 1),
         rebound_goal = ifelse(is.na(rebound_goal), 0, 1),
         odd_man = paste0(substr(odd_man, 1, 1), "-",
                                    substr(odd_man, nchar(odd_man), nchar(odd_man)))
  )
```

```{r}
# Change NA-NA to NA in odd_man column for less typing
pass[["odd_man"]] <- ifelse(pass[["odd_man"]] == "NA-NA", "NA", pass[["odd_man"]])
```

```{r}
# Team logos will be drawn on my graphs using the ggimage package.
team_logo <- c("http://www.capsinfo.com/images/NHL_Team_Logos/NHL_Ducks_Primary.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/NHL_Coyotes_Primary.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/NHL_Bruins_Primary.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/NHL_Sabres_Primary.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/carolina.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/NHL_BlueJackets_Primary.png",
                "http://www.capsinfo.com/images/NHL_Team_Logos/calgary.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/chicago.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/colorado.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/NHL_Stars_Primary.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/detroit.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/NHL_Oilers_Primary.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/NHL_Panthers_Primary.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/NHL_Kings_Primary.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/NHL_Wild_Primary.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/montreal.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/newjersey.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/NHL_Predators_Primary.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/NY-Islanders-Primary.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/newyorkr.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/NHL_Senators_Primary.png",
               "http://www.stickpng.com/assets/images/5a4fbba3da2b4f099b95da1a.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/NHL_Penguins_Primary.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/NHL_Sharks_Primary.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/stlouis.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/NHL_Lightning_Primary.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/NHL_MapleLeafs_Primary.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/Vancouver_Canucks.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/NHL_Jets_Primary.png",
               "http://www.capsinfo.com/images/NHL_Team_Logos/NHL_Capitals_Primary.png"
)

```

### Examine the dataset
```{r}
pass %>% 
  select(shot_on_goal, goal, odd_man, shot_type, a1_zone, rebound_shot_on_goal) %>% 
  head(10) %>% 
  kable() %>% 
  kable_styling(bootstrap_options = c("striped", "hover"))
```

There are 25 variables and over 48,000 observations in this dataset. Some notable variables that I will be using for my analysis include shot_on_goal (1 for yes and 0 for no), goal (1 for yes and 0 for no), odd_man (number of players involved in odd man rushes), shooting_percentage (number of goals / number of shots on goals), shot_type (one-timer, slap shot, wrist/snap shot, backhand, etc), a1_zone (zone that the primary assist originated), and rebound_shot_on_goal (if a rebound occured and resulted in a shot on goal).

There are a couple literature that analyze this dataset. [Ryan Stimson's paper](http://www.sloansportsconference.com/wp-content/uploads/2017/02/1614.pdf) discusses repeatability of certain metrics on a team level and player level. This post on [NHLNumbers](https://www.nhlnumbers.com/2016/08/10/passing-project-dangerous-primary-shot-contributions) discusses repeatability of royal road passing and nicely provides visualizations. 

In this blog post, I will solely look at repeatability of number of odd-man rush shots, shooting percentages, and rebound shooting percentages in 5 on 5 situations.



### Repeatability of odd-man rush shots

First, let's examine if shots taken in odd-man rushes are repeatable. I measured repeatability by using **split half correlations**. I split the 2015-2016 season into two halves by designating the midpoint of the season as Jan 1st, 2016. All the games before that date are labelled as the first half of the season while the games after that date are labelled as the second half.

```{r, warning=FALSE}
pass_odd_man <- pass %>%
  # Indicate whether data is from first half of the season (Before 2016-01-01) or the second half (After 2016-01-01)
  mutate(half = if_else(date > as.Date("2016-01-01"), "second", "first")) %>% 
  # Further divide odd_man rushes by number of advantage in men
  mutate(odd_man = ifelse(odd_man == "NA-NA", "non_odd_man",
                          ifelse(odd_man %in% c("1-0", "2-1", "3-2", "4-3"), "one_man", 
                                 ifelse(odd_man %in% c("2-0", "3-1", "4-2", "3-0"), "two_plus_man",
                                        "NA")))) %>% 
  filter(strength == "5v5") %>% 
  group_by(team, half, odd_man) %>%
  # Get total number of shots for each team, half, odd man situation
  summarise(shots = n()) %>% 
  filter(!is.na(team), !is.na(half)) %>%
  spread(key = half, value = shots, fill = 0) %>% 
  ungroup()

```

```{r}
### Note:At this point, I realized that the number of rows of teams did not match the number of rows of urls of team logos, so I decided to create a function

# Create a function to match number of rows of team logos urls to number of rows for each team
match_logo <- function(df) {
  # Configure team logos to match values in odd_man
  team_freq <- as.data.frame(table(df[["team"]]))
  empty_list <- list()
  for (ii in seq(nrow(team_freq))) {
    empty_list[[ii]]<- rep(team_logo[ii], times =  team_freq[,2][ii])
  }
  
  return(as.vector(unlist(empty_list)))
}
```


```{r, fig.width = 9, fig.height = 4.5}
# Let's add on team logo urls to pass_odd_man
pass_odd_man <- pass_odd_man %>% 
  mutate(team_logo_url = match_logo(pass_odd_man))

# Let's change the labels on the facets:
odd_man <- list(
  "NA" = "No player advantage",
  "one_man" = "One player advantage",
  "two_plus_man" = "Two+ player advantage"
)

odd_man_labeller <- function(variable,value){
  return(odd_man[value])
}


# Plot
ggplot(pass_odd_man, aes(x = first, y = second)) +
  geom_image(aes(image = team_logo_url), size = 0.08) +
  geom_smooth(method = "lm") + 
  facet_wrap(~odd_man, scales = "free", labeller = odd_man_labeller) +
  labs(x = "First Half Shots",
       y = "Second Half Shots") +
  ggtitle("Repeatability of Total Shots Facetted by Odd Man Rushes") +
  coord_fixed()
```

_I can assert repeatability in linear regression graphs used in this article when I can NOT draw a horizontal line through the confidence band surrounding the regression line. This reveals that there does exist a significant relationship between the past (First Half Shots) and the future (Second Half Shots), which indicates repeatability of shots._

This facetted graph shows that shots taken on odd-man rushes are repeatable over a season when there is a one player advantage and no player advantage. This leads me to think that its increadibly hard to predict the number of shots in two-plus player advantages over a season. Perhaps, teams with an incredible advantage in rushes have too many options at their disposal, and thus these rushes do not always result in shots on net. Rather, they could make one extra pass that extinguishes the offensive threat or the extra pass could be intercepted by the defender(s). As there is no repeatability in two plus player advantage, we will ignore these situations for this analysis.


### Repeatability of shooting percentage

Let's examine shooting percentages and see if we can detect the same trend in the upcoming facetted graphs.

```{r, fig.width = 9, fig.height = 4.5}
pass_shooting_percentage <- pass %>%
  mutate(half = if_else(date > as.Date("2016-01-01"), "second", "first")) %>% 
  mutate(odd_man = ifelse(odd_man == "NA-NA", "non_odd_man",
                          ifelse(odd_man %in% c("1-0", "2-1", "3-2", "4-3"), "one_man", 
                                 ifelse(odd_man %in% c("2-0", "3-1", "4-2", "3-0"), "two_plus_man",
                                        "NA")))) %>% 
  filter(strength == "5v5") %>% 
  group_by(team, half, odd_man) %>% 
  # Calculate the shooting percentage, defined as the number of goals over the number of shots on goal
  summarise(shooting_percentage = 100 * (sum(goal) / sum(shot_on_goal))) %>% 
  filter(!is.na(team), !is.na(half)) %>%
  spread(key = half, value = shooting_percentage) %>% 
  ungroup() %>% 
  mutate(team_logo_url = match_logo(.))

# Ignore two_plus_man
pass_shooting_percentage %>% 
  filter(odd_man != "two_plus_man") %>% 
ggplot(aes(x = first, y = second)) +
  geom_image(aes(image = team_logo_url), size = 0.06)+
  geom_smooth(method = "lm") +
  facet_wrap(~odd_man, scales = "free", labeller = odd_man_labeller) +
  coord_fixed() +
  labs(x = "First Half Shooting %",
       y = "Second Half Shooting %") +
  ggtitle("Repeatability of Shooting % Facetted by Odd Man Rushes")
```

The scales for the two facetted graphs are vastly different: shooting percentages across the whole season are _significantly higher_ in "one player advantage" situations, which is reasonable given the offensive freedom during such rushes. Interestingly, shooting percentages are indeed not repeatable over a season during "no player advantage" rushes since I can draw a horizontal graph through the confidence band. In most situations, shooting percentages are not repeatable, but when there is a one-man advantage, they are repeatable. I figure that its alot more easier to predict shooting percentages in one man rushes because they present the most limited options: _either pass to a teammate or shoot at the net._ This consistency in offensive options may lead to more repeatability in shooting percentages.



### Shooting percentages by Zone of Primary Pass

Now, I observe the difference in repeatability of shooting percentages during odd man rushes according to the zone of the primary pass.

```{r, warning=FALSE, fig.width = 9, fig.height = 4.5}
pass_shooting_percentage_zone <- pass %>%
  mutate(half = if_else(date > as.Date("2016-01-01"), "second", "first")) %>% 
  # Extract out the last letter in a1_zone, which shows the lane in which the pass originiates from
  mutate(a1_zone = substr(pass$a1_zone, start = nchar(pass$a1_zone), stop = nchar(pass$a1_zone)),
         a1_zone = tolower(a1_zone),
         a1_zone = ifelse(a1_zone == "c", "center_lane",
                          ifelse(a1_zone == "f", "faceoff",
                                 ifelse(a1_zone == "l", "left_lane",
                                        ifelse(a1_zone == "p", "point",
                                               ifelse(a1_zone == "r", "right_lane", NA)))))) %>% 
  filter(strength == "5v5", odd_man != "NA") %>% 
  group_by(team, half, a1_zone) %>% 
  summarise(shooting_percentage = 100 *  (sum(goal) / sum(shot_on_goal))) %>% 
  filter(!is.na(team), !is.na(half), !is.na(a1_zone)) %>% 
  spread(key = half, value = shooting_percentage) %>% 
  ungroup() %>% 
  mutate(team_logo_url = match_logo(.))




# Let's change the labels on the facets:
pass_lanes <- list(
  "left_lane" = "Left Lane",
  "faceoff" = "Faceoff",
  "center_lane" = "Center Lane",
  "point" = "Passes back to the point",
  "right_lane" = "Right Lane",
  "not_categorized" = "None"
)

pass_lanes_labeller <- function(variable,value){
  return(pass_lanes[value])
}

# Change order of facets
pass_shooting_percentage_zone$a1_zone <- factor(pass_shooting_percentage_zone$a1_zone,
                                                levels = c("left_lane",
                                                           "faceoff",
                                                           "center_lane",
                                                           "point",
                                                           "right_lane")) 

# Plot
 pass_shooting_percentage_zone %>%
   filter(a1_zone %in% c("left_lane", "center_lane", "right_lane")) %>% 
   ggplot(aes(x = first, y = second)) +
   geom_image(aes(image = team_logo_url), size = 0.09)+
   geom_smooth(method = "lm") +
   facet_wrap(~a1_zone, scales = "free", labeller = pass_lanes_labeller) +
   coord_fixed() +
   labs(x = "First Half Shooting %",
        y = "Second Half Shooting %") +
   ggtitle("Repeatability of Shooting % During Odd Man Rushes Facetted By Lanes of Passes")
```

None of these graphs show any signs of repeatability since I can draw a horizontal line through all these graphs. As a result, the lane of a primary pass does not help explain the discrepancy in repeatability of shooting percentages in odd man rushes.

### Repeatability of rebound shots

During the course of a game, there is bound to be a rebound. Most "trash" goals scored in the "dirty" areas are scored off of a rebound. Let's examine the repeatability of rebound shots over a season.

```{r}
pass %>% 
  mutate(half = if_else(date > as.Date("2016-01-01"), "second", "first")) %>% 
  filter(strength == "5v5") %>% 
  group_by(team, half) %>% 
  summarise(rebound_shot = sum(rebound_shot_on_goal)) %>% 
  filter(!is.na(team), !is.na(half)) %>% 
  spread(key = half, value = rebound_shot) %>% 
  ungroup() %>% 
  ggplot(aes(x = first, y = second)) +
  geom_image(aes(image = team_logo), size = 0.04, by = "height") +
  geom_smooth(method = "lm") +
  coord_fixed() +
  labs(x = "First Half Rebound Shots",
       y = "Second Half Rebound Shots") +
  ggtitle("Repeatability of Rebound Shots")
```

This graph clearly shows that rebound shots are repeatable over a given season. This suggests that teams have different offensive strategies and different players with different shooting styles that make the occurence of rebound shots predictable over a given season. For example, the Pittsburgh Penguins, Washington Capitals, and Tampa Bay Lightning have some of the best shooters in the game in Kessel, Ovechkin, Kucherov, and Stamkos. These players, who take many shots, must significantly contribute to their respective teams' rebound percentage over a season. If they have high shooting percentages, this may lead to lower rebound percentages (shots that end up in the net do not count as rebounds) and if they have low shooting percentages, this may lead to higher rebound percentages. 


### Repeatability of rebound shooting percentage

To dig deeper into the effectiveness of rebounds in offensive situations, I've defined _rebound shooting percentage as the percentage of rebound goals scored._ Let's now examine its repeatability.

```{r, fig.width = 9, fig.height = 4.5}
pass %>% 
  mutate(half = if_else(date > as.Date("2016-01-01"), "second", "first")) %>% 
  filter(strength == "5v5") %>% 
  group_by(team, half) %>% 
  summarise(rebound_percentage = 100 * (sum(rebound_goal) / sum(rebound_shot_on_goal))) %>% 
  filter(!is.na(team), !is.na(half)) %>% 
  spread(key = half, value = rebound_percentage) %>% 
  ungroup() %>% 
  ggplot(aes(x = first, y = second)) +
  geom_image(aes(image = team_logo), size = 0.04, by = "height") +
  geom_smooth(method = "lm") +
  coord_fixed() +
  labs(x = "First Half Rebound %",
       y = "Second Half Rebound %") +
  ggtitle("Repeatability of Rebound Shooting %")
```

We see that overall, rebound percentages are not repeatable since I can draw a horizontal line through the confidence band. This suggests that rebounds are a random occurence during a hockey game. Perhaps, different goaltenders have varying abilities in controlling rebounds. Furthermore, over an entire season, there exist random offensive situations that may cause rebounds and thus, make it impossible to predict such a statistic. 

## Conclusion

After observing these four different graphs, I found that shots taken during no player advantages and one player advantages are repeatable; shooting percentages are only repeatable during one player advantages; No zones of primary passes reveal any discrepancy in shooting percentages; rebound shots are indeed repeatable but rebound shooting percentages are not repeatable. 

